---
title: "Midterm Project"
author: "Mylène Feuillade and Gustave Kenedi"
output:
    html_document:
        df_print: paged
        toc: true
        toc_depth: 3
        toc_float:
            collapsed: false
        theme: united
        highlight: textmate
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
    include = TRUE,
    message = FALSE,
    warning = FALSE,
    cache = FALSE
)
```

## Logistics

-   ***DEADLINE***: The midterm is due before our class on 26 of
    October.
-   You can do this homework in groups of **two at most**.
-   This homework has a total of 20 points. The number of points for
    each section is clearly indicated in the section and/or subsection
    headings.
-   You need to hand in this `.Rmd` file on Moodle and send me the
    `.html` file - that is produced when you click on the "Knit"
    button - through Slack. If you are having issues knitting your file
    do get in touch (and **knit regularly to make sure everything is
    working properly**).
-   We left code blocks in this `.Rmd` file for you to fill out. Just
    put your code within the marked region, and click the *Knit* button
    in RStudio, just above this script, to see the generated HTML
    output. Knit it now, before you add any code to see how nice the
    output looks like.
-   Whenever we give you a chunk of code with code already written, you
    need to replace `eval = FALSE` by `eval = TRUE` in the chunk
    "header". This will make sure that the chunk is 'evaluated' (i.e.,
    that the code runs) when you knit the document.
-   FYI a code block looks like this:

```{r qx.x, echo = TRUE, eval = TRUE}
# code goes here
```

-   **Do NOT delete the \`\`\`{r} and \`\`\`**; they separate the code
    chunk from text.
-   You can find a nice intro to Rmarkdown (`.Rmd`)
    [here](https://rmarkdown.rstudio.com/authoring_quick_tour.html).
-   Questions marked with a red star ([\*]{style="color: red;"}) are the
    ones we will build on progressively to clean the data until it is
    ready for the regression analysis. Other questions are side quests,
    which you don't need to answer to get to the regressions.
-   We have added a link to a clean dataset just before the regression
    analysis part. If there are necessary questions you have not managed
    to do, you can start from this clean data instead of using your own.

# 2020 US Presidential Elections - Analysis of the Results

This midterm project consists in one long analysis of data from the 2020
Presidential elections in the United States.

You will be asked to download the data from its original source, just
like you would if you were to undertake this task by yourself. You will
use everything we've learned so far: importing (un-tidy) data;
summarizing, visualizing and tidying it; running regressions for
different variables; and thinking about whether these associations can
be interpreted causally. In the process, you will also learn 2 useful
operations: merging 2 (or more) datasets together, and "reshaping" data
from wide to long format. Don't worry we will work you through how to do
all of this.

We hope you find this project interesting and useful. Now let's get
going!

**Please add your names here**

## Loading packages

It's good practice to load all the packages you will need in the same
place.

```{r load packages}
# Load all the packages you need here
# Don't write the install.packages() code here. Run that code in the console.
library(readxl)
library(readr)
library(tidyverse)
library(janitor)
library(sf)
library(tmap)
library(skimr)
library(jtools)
library(huxtable)
library(rmapshaper)
```

## Loading the Data [0.5 points]

Download the data from November 2020 US presidential elections from
here:

> <https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ#>

Scroll down, click on the download icon next to the second file
(countypres_2000-2020.tab), and select "Comma Separated Values (Original
File Format)".

This is straight from the MIT Election Data + Science lab, which
collects and shares election data from the United States. This dataset
contains election results aggregated at the county level.
[Counties](https://en.wikipedia.org/wiki/County_(United_States)) are
small geographic units in the United States.

1[\*]{style="color: red;"}. Load the file you downloaded into RStudio in
an object called `elections`. You can use the `read.csv` function.

```{r q1.1, echo = TRUE, eval = TRUE}
elections <- read.csv(file = './countypres_2000-2020.csv');
```

## Understanding what the dataset contains [1 point]

1\. Use `head` to display the first 6 rows of the dataset

```{r q2.1, echo = TRUE, eval = TRUE}
head(elections, n = 6L);
```

2\. What years are included in our data? What's inside the enigmatic
"mode" variable? Tabulate the variables `year` and `mode` separately to
find out.

```{r q2.2, echo = TRUE, eval = TRUE}
year <- elections[, c("year")];
mode <- elections[, c("mode")];
year <- unique(year, incomparables = FALSE);
mode <- unique(mode, incomparables = FALSE);
```

**Year 2000, 2004, 2008, 2012, 2016, 2020 are included. the "mode"
variable seems indicate the types of vote counted in each row.**

3\. What is the unit of observation? In other words, what does each row
correspond to?

**Each row represents the total number of a specific type of votes that
one US President candidate receive in a specific combination of county +
state + year**

4\. How many variables are there?

```{r q2.4, echo = TRUE, eval = TRUE}
ncol(elections);
```

**There are 12 variables**

5\. How many rows are there?

```{r q2.5, echo = TRUE, eval = TRUE}
nrow(elections);
```

**There are 72617 rows**

## Filtering the data and exploring further [1 points]

1[\*]{style="color: red;"}. This data contains the results of all
presidential elections from 2000 to 2020. We only want to analyse the
2020 results. Use the `filter` function to keep only the observations
that concern the 2020 election. Create a new object `elections_2020` for
this task.

```{r q3.1, echo = TRUE, eval = TRUE}
elections_2020 <- filter(elections, year == 2020);
```

2\. Our main variable of interest is `candidatevotes`. In how many
counties did a candidate get zero votes? How many different candidates
were concerned?

```{r q3.2, echo = TRUE, eval = TRUE}
zero_votes <- filter(elections_2020, candidatevotes == 0);
counties <- zero_votes[, c("county_name")];
counties <- unique(counties, incomparables = FALSE);
nrow(table(counties));
zero_votes_candidates <- zero_votes[, c("candidate")];
zero_votes_candidates <- unique(zero_votes_candidates, incomparables = FALSE);
zero_votes_candidates;

```

**There are 549 counties where a candidate got 0 votes. There are 3
named candidate that got 0 votes in at least 1 county: JO JORGENSEN,
JOSEPH R BIDEN JR, and DONALD J TRUMP. The other candidates concerned
were unnamed**

3\. We saw previously that there were different modes of voting recorded
in the data, and that one category is "Total". This could be worrying:
are we counting ballots twice? Could a ballot appear both in the "in
person" and "total" categories? Cross-tabulate the `mode` variable with
`state`, to understand whether the modes of voting are the same in all
States. (Use the `table` function)

```{r q3.4, echo = TRUE, eval = TRUE}
mode_state <- table(elections_2020$mode, elections_2020$state);
mode_state;
```

As you can see, in some States the `mode` variable takes exclusively the
value "TOTAL" (in Alabama for instance). In those States, we know the
total number of votes obtained by candidates in each county. In other
States, we do not have the total number of votes, but only a number
disaggregated by voting method (in Arizona for instance).

All of those situations are fine, since they mean ballots are not
counted twice.

However there is one worrying case: Utah. In this State, we observe both
the total number of votes, and the number of votes by mail, election day
voting, and early voting. By looking more closely at the data, you can
see that in the Salt Lake county (code 49035), we have the number of
votes received early; on election day; and by mail. For all other
counties, the ballot count for these voting modes is zero, and we only
observe the "total" ballot count. If you wanted to be convinced more
formally that this is true, you could run the following script:

```{r q3.5, echo = TRUE, eval = FALSE}
elections_2020 %>% 
  filter(state=="UTAH") %>% #keep only UTAH observations
  mutate(mode_vote_total = (mode=="TOTAL" & candidatevotes!=0), #create a dummy if we observe a ballot in the "total" category
         mode_vote_other = (mode!="TOTAL" & candidatevotes!=0)) %>% # create a dummy if we observe a ballot in any other voting category
  group_by(county_fips, county_name) %>%  # group by county
  summarise(mode_vote_total = max(mode_vote_total), #create a dummy if for any candidate in the county we observe a ballot in the "total" category
            mode_vote_other = max(mode_vote_other)) %>% #create a dummy if for any candidate in the county we observe a ballot in any other voting category
  View() #visualise the resulting dataset

#You can see that in no county do we observe a ballot both in "total" and in another category!
```

So it holds true that we never count a ballot twice: in every county,
either we count it in the "total" category, or we count it in one of the
categories associated to specific voting method. This is good news!

## Checking quality of the data [2 points]

1\. In States like Arizona where we don't have the total number of
ballots, it is not clear that the voting modes for which we have data
cover all voting methods available.

For instance, in Kentucky we only have the number of ballots cast on
election day. A quick search online shows that early voting is possible
in this State, although it is heavily restricted. It is hence very
likely that some ballots are missing in our data.

Now that we noticed this feature of the data, we want to check how big
of an issue this is: how many votes are not counted in our data? Can we
still reasonably use it, or is its quality too bad?

To answer this question, we will compare the total number of votes by
State in our dataset to the official number provided by the Federal
Government.

Create a new object called `elections_2020_state_totals`, with two
variables: `state_po` and `sum_votes_state`: the sum of `candidatevotes`
by State. Hint: you should group your data by `state_po`, and then use
`summarise`. You should also disregard missing values when you compute
the sum of `candidatevotes`: find out the argument of the `sum` function
which lets you do that.

```{r q4.1, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_state_totals <- 
  elections_2020 %>% 
    group_by(state_po) %>%
    summarise(sum_votes_state = sum(candidatevotes, na.rm = TRUE), .groups = "drop");
elections_2020_state_totals <- unique(elections_2020_state_totals);
```

2\. The official election data in the US is provided by the Federal
Election Commission, in [this
document](https://www.fec.gov/resources/cms-content/documents/2020presgeresults.pdf).
Notice that it is in pdf format. We already extracted it into an excel
file for you.

> Download the file
> [here](https://www.dropbox.com/s/nnzu2i0xmpovqt0/state_votes.xlsx?dl=1)

Import it into a new object named `official_state_votes`, using the
`read_excel` function from the `readxl` package.

```{r q4.2, echo = TRUE, eval = TRUE}
# code goes here
official_state_votes <- read_excel('./state_votes.xlsx');
```

3\. Both your `elections_2020_state_totals` and `official_state_votes`
should have 51 observations, one per State in the US + Washington DC.
But now the official number of votes and the one from our dataset are in
two different R objects. You could compare the numbers manually, but
this would be a lot of work. Instead, we want to merge them into a
single object.

We will do this using the `left_join()` function from the `tidyverse`
package.

This is how `left_join()` works: `left_join(x,y)` will return all rows
from `x` (very important!) and all columns from `x` and `y`. Rows in `x`
with no match in `y` will have `NA` values in the new columns. Rows in
`y` with no match in `x` will be discarded.

The following cheatsheet might help you visualise the way the function
works:
<https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf>.
You can also find more details
[here](https://dplyr.tidyverse.org/reference/mutate-joins.html).

If nothing is specified, `left_join` will automatically detect the
variables which have the same name in both datasets and use them to join
the data. If your variables have different names in the datasets, you
can specify
`left_join(x,y, by=c("var1x" = "var1y", "var2x" = "var2y"))`. This will
match variable `var1x` in `x` to `var1y` in `y` and `var2x` in `x` to
`var2y` in `y`. (You can keep going, and join on more than two
variables!)

Create a new object called `elections_2020_state_comparison`, which is
the `left_join` of `elections_2020_state_totals` (1st argument) and
`official_state_votes` (2nd argument). Here, our identifying variables
have different names, so we need to add the argument
`by=c("state_po" = "state_short")`

```{r q4.3, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_state_comparison <- left_join(elections_2020_state_totals, official_state_votes, by = c("state_po" = "state_short"));
```

Congratulations, you have merged your first datasets! Easy right? :)

4\. Compute the percentage difference between the official number of
votes, and the sum of votes in our data by State:
`(official votes - sum votes)/official votes*100`. What are the min, max
and mean of this percentage difference?

```{r q4.4, echo = TRUE, eval = TRUE}
# code goes here
percentageDifference <- (elections_2020_state_comparison$official_total_votes - elections_2020_state_comparison$sum_votes_state) / elections_2020_state_comparison$official_total_votes * 100;
min(percentageDifference);
max(percentageDifference);
mean(percentageDifference);
```

**The min is -0.925%, the max is 0.545%, the mean is -0.044%**

If our election data contained the official number of votes per State,
the difference would always be zero. Clearly, this is not the case: in
some States our data has too few votes, in others it has too many.
However, the difference is always very very small and should not
dramatically change the final results obtained by each candidate.

Still, the discrepancy is not easy to explain: the MIT lab collects
official results from all the States and puts them together into a
national file. It is not clear where the difference is coming from:
maybe there was human error in data collection, maybe the results were
updated in some States after a recount...

Since the United States government itself does not provide official
county-level results for the entire country, the only alternative to
using the MIT data would be to start from State level official results.
However, this would be a lot of work as we would have to find all the
files, deal with format issues (many data files are still provided in
pdf, for instance [in
Kentucky](https://elect.ky.gov/results/2020-2029/Documents/2020%20General%20Election%20Results.pdf))
and clean them all. It would be time consuming, and we would likely make
more errors than the MIT lab!

## Some more tidying [2.5 points]

1\. Now that we have decided that our data was good enough, we still
need to tidy it before we can use it. Make sure that the candidate name
is always written with the same format by tabulating the `candidate`
variable.

```{r q5.1, echo = TRUE, eval = TRUE}
# code goes here
table(elections_2020$candidate);
```

2[\*]{style="color: red;"}. Remember that we noticed earlier there were
missing values in the `candidatevotes` and `totalvotes` variables.

Run the following script to fix the issue (replace `eval = FALSE` by
`eval = TRUE` below, so that the code is also run when you knit). This
chunk of code:

-   replaces `candidate_votes` by zero when it is missing
-   replaces `total_votes_county` by the sum of votes in the county when
    it is missing.

Note that we create a new object, `elections_2020_clean`.

The `ifelse` function works in the following way:
`ifelse(test, value if test==TRUE, value if test==FALSE)`. For instance
in the third line below, for each observation, if `candidatevotes` is
missing then it is replaced by 0, otherwise it is not changed.

```{r q5.2, echo = TRUE, eval = TRUE}
elections_2020_clean <- elections_2020 %>%
  group_by(county_fips) %>%
  mutate(candidatevotes = ifelse(is.na(candidatevotes), 0, candidatevotes),
         totalvotes = ifelse(is.na(totalvotes), sum(candidatevotes), totalvotes))
```

3[\*]{style="color: red;"}. Clean some more the `elections_2020_clean`
object by:

-   Summarising the data so that the unit of observation is a candidate
    in a county. To do this, you need to sum the number of votes
    obtained for each candidate in each county, across all modes of
    voting. Note: in order to keep all the variables, you will need to
    use the following command
    `group_by(state, state_po, county_name, county_fips, totalvotes, candidate, party)`.
    If we only grouped by `county_fips, candidate`, we would "lose" all
    the other "extra" variables that give us information about the
    county (its name, its state, the number of votes, ...). If this is
    not clear to you, simply run the code twice, once grouping by all
    the variables, and once with `group_by(county_fips, candidate)`, and
    see what happens!

-   Creating a variable `pct_votes` which is equal to the percentage of
    votes obtained by a candidate in a county: (number of votes obtained
    by a candidate / total number of votes in county)\*100.

-   Grouping the data by county (`county_fips`), and then use a
    combination of `rank()` and `desc()` commands to compute the ranks
    of each candidate within each county. You should create a variable
    called `candidate_rank` which takes the value 1 for the candidate
    who received the most vote in the county, 2 for the candidate who
    received the second highest number of votes etc.

-   Once you've ensured your code is correct, "ungroup" the data by
    passing your object into the `ungroup()` function or by piping it.
    This function does what it says: it removes grouping. This means
    that it cancels the effect of your last `group_by`, and ensures that
    the next functions you use will not be applied by group.

You can pipe all these commands together if you wish or proceed in
several steps.

```{r q5.3, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_clean <- elections_2020_clean %>%
  group_by(state, state_po, county_name, county_fips, totalvotes, candidate, party) %>%
  summarise(candidatevotes = sum(candidatevotes), .groups = "drop");

 elections_2020_clean <- elections_2020_clean %>%
   group_by(county_fips) %>%
   summarise(candidate = candidate, party = party, state = state, state_po = state_po, county_name = county_name, candidatevotes = candidatevotes, totalvotes = totalvotes, pct_votes = (candidatevotes / totalvotes) * 100, candidate_rank = rank(desc(candidatevotes)), .groups = "drop") %>%
   ungroup();
```

4\. Check that your variable `pct_votes` is actually a percentage (that
it is comprised within 0 and 100). This is a safety check to make sure
that there is no inconsistency in the data.

```{r q5.4, echo = TRUE, eval = TRUE}
# code goes here
min(elections_2020_clean$pct_votes) >= 0;
max(elections_2020_clean$pct_votes) <= 100;
```

## Maps [3 points]

Election data are probably better visualized using maps.

In this section we will produce two maps displaying:

1.  Joe Biden and Donald Trump's vote percentages in each county,
2.  The leading candidate in each county.

To produce the maps you will need to install and load the `sf`, `tmap`
and `rmapshaper` packages.

Maps are tricky and very complicated objects. But to boil it down to the
essential ingredients, you need a "shapefile" which is basically a file
that contains the map onto which data will be plotted.

This file is provided by the US census website
[here](https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_county_500k.zip).
However it is somewhat heavy, so to make your life easier, we have
simplified it for you using the following commands. **Do NOT** touch nor
run this chunk, and **do NOT** replace `eval = FALSE` by `eval = TRUE`.
We've left it so you can see how we created the simplified shapefile you
will be loading in the next question.

```{r q6.0 simplifying code, echo = TRUE, eval = FALSE}
county_shfl_full <- read_sf("path/cb_2018_us_county_500k/cb_2018_us_county_500k.shp")
# where `path` corresponds to where the `cb_2018_us_county_500k` folder is located.
county_shfl <- st_simplify(county_shfl_full, preserveTopology = FALSE, dTolerance = 1000) #simplify the map to make it less memory-heavy

save(county_shfl, file="path/county_shfl.RData") #saving the object
```

You can download the resulting R object from this link: \>
<https://www.dropbox.com/s/m8zq5e38imp2he1/county_shfl.RData?dl=1>

1\. Let's load the `county_shfl.RData` file by using the `load()`
function. Replace `eval = FALSE` by `eval = TRUE`.

```{r q6.1, echo = TRUE, eval = TRUE}
load("./county_shfl.RData")
# where `path` corresponds to where the county_shfl.RData file is located on your laptop.
```

You should now have an object called `county_shfl` in your environment

2\. FIPS codes are numbers which uniquely identify geographic areas in
the US. Working with FIPS codes is always advisable when identifying
areas, as names can be duplicates: there is an Aidar county in Kentucky
and another one in Iowa.

What is the county FIPS code of the "MCDUFFIE" county in the
`elections_2020_clean` data.frame? What are the State FIPS (`STATEFP`),
county FIPS (`COUNTYFP`) and `GEOID` of the same county in the
`county_shfl` data? Hint: in `county_shfl` the county's name is written
"McDuffie" and the variable containing county names is `NAME`.

```{r q6.2, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_clean[elections_2020_clean$county_name == "MCDUFFIE",];
county_shfl[county_shfl$NAME == "McDuffie",];
```

**The county FIPS if McDuffie in elections_2020_clean is 13189. In
county_shfl, the State FIPS is 13, County FIPS is 189, and the GEOID is
13189 for McDuffie**

3\. You might have noticed that the `county_fips` variable in our
`elections_2020_clean` is actually made of the State FIPS code, followed
by the county code.

In `county_shfl`, we have a variable for the State FIPS code and a
separate one for the county code. The variable containing the same code
as in `elections_2020_clean` is actually called `GEOID`. This type of
thing is exactly why you always need to look inside your data: it would
not have been obvious from the variable names.

What is the type of `county_fips` in the `elections_2020_clean` data?
What is the type of `GEOID` in `county_shfl`? Hint: use the `typeof`
function.

```{r q6.3, echo = TRUE, eval = TRUE}
# code goes here
typeof(elections_2020_clean$county_fips);
typeof(county_shfl$GEOID);
```

**The type of "county_fips" is integer while the type of "GEOID" is
character**

It is not the same! In order to merge the two datasets, we want
`county_fips` to have the same type as in `GEOID` and be a character
string instead.

Furthermore, because it is stored as a numeric value there is no leading
zero (the FIPS code "01001" is stored as 1001).

The following script fixes this, and creates a new variable called
`county_fips_clean`. Run the code and change `eval = FALSE` to
`eval = TRUE` to make sure the code runs when you knit your document.

We start by using the `as.character` function to convert the variable to
a character string. Then, we use the `ifelse` function again. What this
conditional statement says is: if the number of characters of the
variable `county_fips_clean` is equal to 4, then add a "0" in front of
this variable; otherwise do nothing.

```{r q6.3 given, echo = TRUE, eval = TRUE}
elections_2020_clean <- elections_2020_clean %>% 
  mutate(county_fips_clean = as.character(county_fips), #convert county_fips to a character string
         county_fips_clean = ifelse(nchar(county_fips_clean) == 4,
                            paste0("0", county_fips_clean),
                            county_fips_clean)) #Add a leading zero when necessary: if there are only four characters to the FIPS code, it means that the leading 0 is missing
```

4\. How many different State FIPS codes are there in the `county_shfl`
dataset? Hint: You can use a combination of the `length` and `unique`
commands to count them.

```{r q6.4, echo = TRUE, eval = TRUE}
# code goes here
length(unique(county_shfl$STATEFP));
```

**There are 56 different State FIPS codes**

5\. Our shapefile clearly contains the map for more than just the 50 US
States + DC. By looking at the existing State codes
[here](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/?cid=nrcs143_013696),
you can see that codes above 56 correspond to overseas territories.

The final map we want to produce should only contain the 50 States + DC:
we want to remove overseas territories. We also exclude Alaska and
Hawaii for purely aesthetic reasons: including them zooms out the map
too much. We could fix this problem by showing Alaska and Hawaii in a
corner of the map but this is beyond the scope of the midterm.

Filter `county_shfl_clean` to keep only observations such that:

-   `STATEFP` is not equal to "02" (Alaska) or "15" (Hawaii)
-   and `as.numeric(STATEFP)` is lower or equal to 56 (note that because
    State FIPS is stored as a character, we have to convert it to a
    numeric before evaluating whether it is below 56)

Create a new object called `county_shfl_clean`.

```{r q6.5 given, echo = TRUE, eval = TRUE}
# code goes here
county_shfl_clean <- county_shfl %>%
  filter(STATEFP != "02" & STATEFP != "15" & as.numeric(STATEFP) <= 56);
```

Furthermore, some counties' FIPS have changed over the years, and they
differ in our two datasets. The following script will fix all these
inconsistencies for you. The `case_when` function is similar to
`ifelse`. Here it modifies the variable `county_fips_clean`, such that :

-   if `county_fips_clean` was equal to "46113", it now takes the value
    "46102"
-   if `state_po` is equal to "DC", `county_fips_clean` now takes the
    value "11001"
-   else `county_fips_clean` keeps the same value.

Remember to replace `eval = FALSE` with `eval = TRUE`.

```{r q6.5 given part 2, echo = TRUE, eval = TRUE}
elections_2020_clean <- elections_2020_clean %>% 
  mutate(county_fips_clean = case_when(
    county_fips_clean == "46113" ~ "46102",
    state_po == "DC" ~ "11001",
    TRUE ~ county_fips_clean))
```

6\. Use `left_join` to merge `county_shfl_clean` (first argument) and
`elections_2020_clean` (second argument). Create a new object called
`elections_2020_map` for this question.

Here, our identifying variables have different names: `GEOID` in
`county_shfl_clean` (first argument) and `county_fips_clean` in
`elections_2020_clean` (second argument). Look at the help file or go
back to question 3 of the "Checking quality" section to see how to
specify the by argument.

```{r q6.6, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_map <- left_join(county_shfl_clean, elections_2020_clean, by = c("GEOID" = "county_fips_clean"));
```

### Map 1: Candidates' vote percentage

7\. Create a heat map of Biden and Trump's percentage of the vote by
county, using the following code (change `eval = FALSE` to `eval = TRUE`
just below so that the html file displays the map):

```{r q6.7 map1 tmap, echo = TRUE, eval = TRUE}
tmap_mode("view") # we want an interactive map
elections_2020_map %>%
             # only keep Trump and Biden
             filter(candidate %in% c("DONALD J TRUMP", "JOSEPH R BIDEN JR")) %>% 
    tm_shape() +
    tm_borders(col="white", lwd = 0.3) + # white and thin (line width) borders
    tm_fill(
        col = "pct_votes",   # variable to be mapped
        title = "% of votes",   # legend title
        id = "county_name",   # information to display when mouse hovers over a departement
        popup.vars = c("Vote %:" = "pct_votes")) + # variable to display in popup window
    tm_facets(by = "candidate") # create one map per selected candidate
```

Wasn't that easy (well, sort of ;) ) ! And now you have 2 comparable
maps of the vote percentages received by Biden and Trump! Hover over the
counties with your mouse. The name of the county will show up. If you
click on a county you will get the vote percentage received by the
candidate. Isn't that cool!

### Map 2: Leading candidate

8\. Create a map displaying the candidate who received the highest
percentage of votes in each county. Try to create this map by yourself
using the code above (and the comments in it). You don't need to create
any new variables. Don't hesitate to use Google (or your favorite search
engine) to help you out. Don't forget to change the title of the legend.

Add the following argument in tm_fill() to have a nice color palette for
your map: `palette = c("red","blue")`.

```{r q6.8 map2, echo = TRUE, eval = TRUE}
# code goes here
tmap_mode("view") # we want an interactive map
elections_2020_map %>%
             # only keep Trump and Biden
             filter(candidate %in% c("DONALD J TRUMP", "JOSEPH R BIDEN JR")) %>% 
    tm_shape() +
    tm_borders(col="white", lwd = 0.3) + # white and thin (line width) borders
    tm_fill(
        palette = c("red","blue"),
        col = "pct_votes",   # variable to be mapped
        title = "% of votes",   # legend title
        id = "county_name",   # information to display when mouse hovers over a departement
        popup.vars = c("Vote %:" = "pct_votes")) + # variable to display in popup window
    tm_facets(by = "candidate") # create one map per selected candidate
```

## Tidying covariate data [2 points]

A covariate is simply the name we give to variables containing
characteristics of our unit of interest (here, characteristic of
counties). A covariate can be of direct interest if it is an independent
variable in the model. It can also be an extra characteristic we add to
the regression to avoid the omitted variable bias which we will discuss
later in the course.

> Download county-level social data from the American Community Survey
> (ACS) here:
> <https://data.census.gov/cedsci/table?q=DP02&g=0100000US%240500000&tid=ACSDP5Y2019.DP02>

Click "DOWNLOAD TABLE" and select the year 2019 for "ACS 5-Year
Estimates Data Profiles" and download. Unzip the folder once downloaded.

> Repeat for ACS economic data:
> <https://data.census.gov/cedsci/table?q=DP03&g=0100000US%240500000&tid=ACSDP5Y2019.DP03>

1[\*]{style="color: red;"}. Import these csv files (the ones with
`data_with_overlays` in their name) into objects called `acs_soc_2019`
and `acs_eco_2019` respectively.

Note: the automatic name of the csv file when you download it includes
the date and time, so the name of your file will be different to mine!
If you work in groups and want the code to run on all your laptops, make
sure that your files have the same name.

```{r q7.1, echo = TRUE, eval = TRUE}
# code goes here
acs_soc_2019 <- read_csv("./ACSDP5Y2019.DP02-Data.csv");
acs_eco_2019 <- read_csv("./ACSDP5Y2019.DP03-Data.csv");
```   

2\. How many variables are there in each dataset?

```{r q7.2, echo = TRUE, eval = TRUE}
# code goes here
ncol(acs_soc_2019);
ncol(acs_eco_2019);
```

**There are 1227 variables in acs_soc_2019, there are 1099 variables in acs_eco_2019**

3[\*]{style="color: red;"}. That's a lot! We need to keep only the
variables we are interested in, and tidy the data a little. For each acs
data frame:

-   Rename the following variables : (`old_name` -\> `new_name`)

1.  In `acs_soc_2019`: `DP02_0062PE` -\> `pct_high_school`
2.  In `acs_eco_2019`: `DP03_0063E` -\> `mean_income`
3.  In `acs_eco_2019`: `DP03_0030PE` -\> `pct_const_maintenance_occ`

The new name tells you what the variable actually contains.
`pct_high_school` is the share of people in the county who only have a
high school degree (no further education). `pct_const_maintenance_occ`
is the share of civilians aged 16 and above who work in the "Natural
resources, construction and maintenance" occupation.

-   Keep only the variables you just renamed, `NAME` and `GEO_ID`

For this task, create two new objects: `acs_soc_2019_clean` and
`acs_eco_2019_clean`.

```{r q7.3, echo = TRUE, eval = TRUE}
# code goes here
acs_soc_2019_clean <- acs_soc_2019 %>%
  rename(pct_high_school = DP02_0062PE) %>%
  summarise(NAME = NAME, GEO_ID = GEO_ID, pct_high_school = pct_high_school);
acs_eco_2019_clean <- acs_eco_2019 %>%
  rename(mean_income = DP03_0063E, pct_const_maintenance_occ = DP03_0030PE) %>%
  summarise(NAME = NAME, GEO_ID = GEO_ID, mean_income = mean_income, pct_const_maintenance_occ = pct_const_maintenance_occ);
```

4\. Display the first five lines of `acs_soc_2019_clean` and
`acs_eco_2019_clean`

```{r q7.4, echo = TRUE, eval = TRUE}
# code goes here
  head(acs_soc_2019_clean, n = 5L);
  head(acs_eco_2019_clean, n = 5L);
```

5\. What is in the first line of the datasets?

**The first line of the datasets contains the titles of each column/variable**

6[\*]{style="color: red;"}. Drop the first row of `acs_soc_2019_clean`
and `acs_eco_2019_clean`. Hint: the negative sign should help you.

```{r q7.6, echo = TRUE, eval = TRUE}
# code goes here
acs_eco_2019_clean <- acs_eco_2019_clean[-c(1),];
acs_soc_2019_clean <- acs_soc_2019_clean[-c(1),];
```

7[\*]{style="color: red;"}. Merge the `acs_soc_2019_clean` and
`acs_eco_2019_clean` together using `left_join`. You don't need to
specify the `by` argument here, since the common variabes `GEO_ID` and
`NAME` should have the same name in both datasets.

Create a new object called `acs_2019_clean`.

```{r q7.7, echo=TRUE, eval=TRUE}
# code goes here
acs_2019_clean = left_join(acs_soc_2019_clean, acs_eco_2019_clean);
```

8\. What is the type of the `pct_const_maintenance_occ`, `mean_income`
and `pct_high_school` variables?

```{r q7.8, echo=TRUE, eval=TRUE}
# code goes here
typeof(acs_2019_clean$pct_const_maintenance_occ);
typeof(acs_2019_clean$mean_income);
typeof(acs_2019_clean$pct_high_school);
```

**All 3 are character type**

9[\*]{style="color: red;"}. In the `acs_2019_clean` dataframe, convert
the `pct_high_school`, `pct_const_maintenance_occ` and `mean_income`
variables to numeric, using the `as.numeric()` function. This function
simply converts character strings to numbers: for example
`as.numeric("17.9") = 17.9`. Hint: to convert the variable, simply use
the `dplyr` verb to create a new variable and give it the same name as
the variable you are modifying.

Ignore the warning about NAs being introduced. They are due to
observations for which `pct_high_school` was already missing (it was
equal to "null")

```{r q7.9, echo=TRUE, eval=TRUE}
# code goes here
acs_2019_clean <- acs_2019_clean %>%
  mutate(pct_high_school = as.numeric(pct_high_school),
         pct_const_maintenance_occ = as.numeric(pct_const_maintenance_occ),
         mean_income = as.numeric(mean_income));
```

10\. What is the `GEO_ID` of the county whose `NAME` is "McDuffie
County, Georgia" in the `acs_2019_clean` dataframe?

```{r q7.10, echo=TRUE, eval=TRUE}
# code goes here
acs_2019_clean[acs_2019_clean$NAME == "McDuffie County, Georgia",]$GEO_ID;
```

11[\*]{style="color: red;"}. The geographical ID of the county is
different to what we had last time! You might have noticed that this
time, the 5-digit FIPS code is preceded by a prefix: "0500000US". We
want to extract the FIPS code from this geographic identifier.

Create a new variable, called `county_fips_clean` which contains only
the 5 last characters of the `GEO_ID` variable. Use the `substr()`
function. This function extracts a substring from a character string.
You need to give it the original string, and the position of the first
and last character of the substring. Example:
`substr("abcde", 2, 4) = "bcd"`, i.e. you get letters 2 to 4. Here we
want to extract characters 10 to 14 of the `GEO_ID` variable. As with
`as.numeric` you have to create the variable using the relevant `dplyr`
verb.

```{r q7.11, echo=TRUE, eval=TRUE}
# code goes here
acs_2019_clean <- acs_2019_clean %>%
  summarise(NAME, GEO_ID, pct_high_school, pct_const_maintenance_occ, mean_income, county_fips_clean = substr(GEO_ID, 10, 14));
```

12[\*]{style="color: red;"}. Now that our covariates data is tidy, we
can finally merge it with the elections data. Merge
`elections_2020_clean` and `acs_2019_clean` using `left_join`. Note:
`elections_2020_clean` needs to be the first argument). Create a new
object called `elections_2020_covariates`.

```{r q7.12, echo=TRUE, eval=TRUE}
# code goes here
elections_2020_covariates <- left_join(elections_2020_clean, acs_2019_clean);
```

## Regression analysis [8 points]

1[\*]{style="color: red;"}. In the regression analysis, we will only
focus on the votes obtained by Donald Trump.

-   Filter `elections_2020_covariates` to keep only the observations
    with the votes for Donald Trump.

-   Keep only the following variables: `state_po`, `county_name`,
    `county_fips_clean`, `pct_votes`, `pct_const_maintenance_occ`,
    `mean_income`, `pct_high_school`.

-   Rename `pct_votes` -\> `pct_votes_trump`.

Create a new object called `elections_2020_reg_trump` for this task.

```{r q8.1, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_reg_trump <- elections_2020_covariates %>%
  filter(candidate == "DONALD J TRUMP") %>%
  summarize(state_po, county_name, county_fips_clean, pct_votes, pct_high_school, pct_const_maintenance_occ, mean_income) %>%
  rename(pct_votes_trump = pct_votes);
```

**Support time:** If you are not confident that your answers to the
necessary red-starred questions are all correct, you can now download a
pre-cleaned dataset
[here](https://www.dropbox.com/s/grxvjyjhdd8upas/elections_2020_reg_trump.RData?dl=1).
This file contains the data you *should* have arrived to after all these
questions. This means that even if you made mistakes before, you can get
a fresh start for the regressions and do them correctly. We will not
penalise you for using this data.

If you choose to use this option, you can load the
`elections_2020_reg_trump.RData` file you just downloaded by using the
`load()` function. Replace `eval = FALSE` by `eval = TRUE`.

```{r q8.support, echo = TRUE, eval = TRUE}
load("./elections_2020_reg_trump.RData")
# where `path` corresponds to where the elections_2020_reg_trump.RData file is located on your laptop.
```

2\. Use the `skim()` function from the `skimr` package to obtain summary
statistics for `mean_income`, `pct_high_school` and
`pct_const_maintenance_occ`.

```{r q8.2, echo = TRUE, eval = TRUE}
# code goes here
skim(elections_2020_reg_trump, mean_income, pct_high_school, pct_const_maintenance_occ)
```

We can see that are some missing values. If you want you can look at
which counties have missing data but otherwise, we'll just ignore them
for this midterm.

3\. To facilitate plotting we will reshape the data in long format. (The
difference between long and wide data is explained in this
[article](https://www.statology.org/long-vs-wide-data/).) We are going
to use the `pivot_longer()` function from the `tidyverse` package to
reshape the data. You can find examples of how the `pivot_longer()`
function works by typing `?pivot_longer` in your console. You can also
find information and examples about `pivot_longer` and its associated
function `pivot_wider`
[here](https://tidyr.tidyverse.org/articles/pivot.html)

For our purposes we will use the following 4 arguments in
`pivot_longer()`:

-   `data = elections_2020_reg_trump` (notice that if you use the `%>%`
    pipe you don't need to specify the data argument);
-   `cols = c("mean_income", "pct_high_school", "pct_const_maintenance_occ")`.
    These are the columns to be reshaped.;
-   `names_to = "covariate"`. This is the name of a new variable to be
    created, where we will store the name of the covariate;
-   `values_to = "value"` This is the name of a new variable to be
    created, where we will store the values of the covariates.

If this is somewhat unclear to you, simply run the code and check how it
differs from the previous dataset. This should clear things up.

Create a new object `elections_2020_plot_trump` for this task. View the
created dataframe to ensure it looks as expected.

```{r q8.3, echo = TRUE, eval = TRUE}
# code goes here
elections_2020_plot_trump <- pivot_longer(
  data = elections_2020_reg_trump,
  cols = c("mean_income", "pct_high_school", "pct_const_maintenance_occ"),
  names_to = "covariate",
  values_to = "value"
)
```

4\. Produce 3 graphs displaying the relationship between the vote
percentage for Donald Trump and our 3 covariates. Each graph should be
on a different row. Make sure that your graph has a meaningful y-axis
label and that the y-axis is between 0 and 100 (use the
`scale_y_continuous()` layer for this). (Hint: use the `facet_wrap`
layer in `ggplot` to make these 3 plots in the same graph. Look at the
`facet_wrap` arguments to ensure each graph is on its own row and the
scales of each plot change freely.)

```{r q8.4, echo = TRUE, eval = TRUE}
# code goes here
ggplot(
  data = elections_2020_plot_trump,
  facet_wrap(align ~ .)
);
```

5\. Add a regression line on these plots using the `geom_smooth` layer
to better see the relationship's sign. Set the argument `method` to
`"lm"` (for linear model) and `se` to `FALSE` (so that it does not
display a confidence interval around the regression line; we'll discuss
this towards the end of the course but you can set the argument to
`TRUE` to see what it does).

```{r q8.5, echo = TRUE, eval = TRUE}
# code goes here

```

6\. From the graphs, is the relationship between the different
covariates and Donald Trump's vote share positive or negative? Rather
strong or rather weak?

7\. Run a simple linear regression of Donald Trump's vote percentage on
each of our covariate variables separately (i.e. 3 simple linear
regressions). For this question you should use
`elections_2020_reg_trump` as the data source.\
Use the `export_summs()` function from the `jtools` package to see the
results from the three regressions side by side.

```{r q8.7, echo = TRUE, eval = TRUE}
# code goes here

```

Don't pay attention to the stars next to the coefficients and to the
number in parenthesis below the coefficients, we'll get to them later on
in the course.

8\. What is the interpretation of the coefficients for `mean_income`,
`pct_high_school` and `pct_const_maintenance_occ`? Be precise.

**Answer goes here.**

9\. Why is the coefficient for mean income so small?

**Answer goes here.**

10\. Why does the number of observations not vary for each regression?

**Answer goes here.**

11\. Which variable's variance explains more of the variance in Donald
Trump's vote?

**Answer goes here.**

12\. Can you interpret these estimates causally? Why not? How else could
you interpret the regression on `pct_high_school`? How would you go
about trying to assess whether mean income or the share of the
Resources/Construction/Maintenance occupation at the county level has a
causal impact on the vote percentage for Donald Trump?

**Answer goes here.**

# Well done!

And that's it for this midterm! We hope you had fun, or at least *some*
fun, doing it and that you improved your `R` skills. We also hope you
got a sense of what "real" data looks like and how it can be tamed by
using the great tools that `R` provides.

::: {.tocify-extend-page data-unique="tocify-extend-page" style="height: 0;"}
:::
