```{r}
library(tidyverse)
```
### Elite University and Income

**1a.**

```{r}
income_univ <- read.csv("elite_income_1.csv", header = TRUE)
x <- nrow(income_univ)

str(income_univ)
```

**1b.**

```{r}
elite_student <- income_univ %>%
  filter(university_attended == "elite university")
proportion <- nrow(elite_student) / x
proportion

```

**1c.**

```{r}
elite_avg_income <- income_univ %>%
  select(university_attended, income) %>%
  filter(university_attended == "elite university") %>%
  summarize(mean = mean(income))
  
nonelite_avg_income <- income_univ %>%
  select(university_attended, income) %>%
  filter(university_attended == "non-elite university") %>%
  summarize(mean = mean(income))


difference <- elite_avg_income$mean - nonelite_avg_income$mean
difference
```




**1d.**

no (reference: https://www.youtube.com/watch?v=RKGw2Lp6Y8I)


#2

```{r}
income_univ_unobs <- read.csv("elite_income_2.csv", header = TRUE)
str(income_univ_unobs)
```

**a.**

```{r}
elite_income <- sum(income_univ_unobs$income_elite_university) / nrow(income_univ_unobs)
nonelite_income <- sum(income_univ_unobs$income_non_elite_university) / nrow(income_univ_unobs)

ATE <- elite_income - nonelite_income
ATE
```
**b.**

yes

**c.**

```{r}
income_univ_unobs <- income_univ_unobs %>%
  mutate(treatment = case_when(university_attended == "elite university" ~ "TRUE",
                               university_attended == "non-elite university" ~ "FALSE"))

head(income_univ_unobs)

```



```{r}
#???
selection_bias <- (difference - ATE) / ATE
selection_bias
```



#3.


```{r}
income_univ_rand <- read.csv("elite_income_3.csv")
str(income_univ_rand)
```

**a.**

# Use university_attended_random to calculate SDO_random
```{r}
income_univ_rand <- income_univ_rand %>%
  mutate(treatment_rand = case_when(income_univ_rand$university_attended_random == "elite university" ~ 1,
                                    income_univ_rand$university_attended_random == "non-elite university" ~ 0))

rand_elite_avg_income <- income_univ_rand %>%
  select(treatment_rand, income_elite_university) %>%
  filter(treatment_rand == 1) %>%
  summarize(mean = mean(income_elite_university))
  
rand_nonelite_avg_income <- income_univ_rand %>%
  select(treatment_rand, income_non_elite_university) %>%
  filter(treatment_rand == 0) %>%
  summarize(mean = mean(income_non_elite_university))


SDO_random <- rand_elite_avg_income$mean - rand_nonelite_avg_income$mean
SDO_random
```

# Use randomly generated treatment_rand variable to calculate SDE_randome
```{r}
income_univ_rand <- income_univ_rand %>%
  mutate(treatment_rand = sample.int(2, nrow(income_univ), replace=TRUE) - 1)

rand_elite_avg_income <- income_univ_rand %>%
  select(treatment_rand, income_elite_university) %>%
  filter(treatment_rand == 1) %>%
  summarize(mean = mean(income_elite_university))
  
rand_nonelite_avg_income <- income_univ_rand %>%
  select(treatment_rand, income_non_elite_university) %>%
  filter(treatment_rand == 0) %>%
  summarize(mean = mean(income_non_elite_university))


SDO_random <- rand_elite_avg_income$mean - rand_nonelite_avg_income$mean
SDO_random
```



**b.**

```{r}
abs((SDO_random - ATE) / ATE)
```

**c.**

2???

#Socmob


```{r}
socmob <- read.csv("socmob_quiz.csv", header = TRUE)
str(socmob)
```


**4a. Counties with missing upward mobility measure**

```{r}
socmob %>%
  filter(is.na(upward_socmob))
```

There are 11 counties with missing upward mobility measure

**4b. Counties with population density of 0**

```{r}
socmob %>% 
  filter(popdensity_2000 == 0)
```

There are 78 counties with population density of 0.

**4c. Filter data to keep counties with non-missing population density greater than 0 and non-missing upward mobility measure**

```{r}
socmob_clean <- socmob %>%
  filter(popdensity_2000 > 0,
        !is.na(upward_socmob))

str(socmob_clean)
```

**5. Create a scatter plot**

```{r}
ggplot(socmob_clean, aes(x = popdensity_2000, y = upward_socmob)) +
  geom_point() 
```

-> Because many counties have a population density of 0.

**6.**

```{r}
socmob_clean <- socmob_clean %>%
  mutate(log_popdensity_2000 = log(popdensity_2000))

ggplot(socmob_clean, aes(x = log_popdensity_2000, y = upward_socmob)) + 
  geom_point(alpha = 0.5)
```

-> There is a nonlinear relationship between the two variables.

**7.**

```{r}
socmob_reg <- lm(upward_socmob ~ log_popdensity_2000, data = socmob_clean)
summary(socmob_reg)
```

-> For each one unit increase in `log_popdensity_2000`, there is a 0.112 percent decrease (?)
-> B

**8.**

-> $R^2$ is 0.1205, so approximately 12% in county-level upward social mobility is explained by the variance in the log of population density.









